\framework: Morphology-based Entity and Relational Entity Extraction Framework for Arabic

### 2018-06-23 tex/merf/TALMajor/abstract.tex
minor changes

Rule-based techniques and tools to extract entities and relational entities from documents allow users to specify desired entities using natural language questions, finite state automata, regular expressions, structured query language statements, or proprietary scripts.
These techniques and tools require expertise in linguistics and programming.
\major{
They lack support of Arabic {\em morphological analysis} which is key to 
process Arabic text.}
In this work, we present \framework; a morphology-based entity and relational entity extraction framework for Arabic text.
\framework provides a friendly interface where the user, with basic knowledge of linguistic features and regular expressions, defines tag types and interactively associates them with regular expressions defined over Boolean formulae.
Boolean formulae range over matches of Arabic morphological features, and synonymity features.
Users define relations with tuples of subexpression matches and can associate code actions with subexpressions.
\framework computes feature matches, regular expression matches, and constructs entities and relational entities from the user-defined relations.
We evaluated our work with several case studies and compared with existing application-specific techniques.
The results show that \framework requires shorter development time and effort compared to existing techniques and produces reasonably accurate results within a reasonable overhead in run time.

>>>>

Rule-based techniques and tools to extract entities and relational entities from documents allow users to specify desired entities using natural language questions, finite state automata, regular expressions, structured query language statements, or proprietary scripts.
These techniques and tools require expertise in linguistics and programming.
They lack support of Arabic morphological analysis which is key to 
process Arabic text.
In this work, we present \framework; a morphology-based entity and relational entity extraction framework for Arabic text.
\framework provides a friendly interface where the user, with basic knowledge of linguistic features and regular expressions, defines tag types and interactively associates them with regular expressions defined over Boolean formulae.
Boolean formulae range over matches of Arabic morphological features, and synonymity features.
Users define relations with tuples of subexpression matches and can associate code actions with subexpressions.
\framework computes feature matches, regular expression matches, and constructs entities and relational entities from the user-defined relations.
We evaluated our work with several case studies and compared with existing application-specific techniques.
The results show that \framework requires shorter development time and effort compared to existing techniques and produces reasonably accurate results within a reasonable overhead in run time.

### 2018-06-23 tex/merf/TALMajor/directionFigure.tex
done till 3.19
format issues ficed: figure and table caption style
minor changes

\transtrue
\setcode{utf8}
\setarab
\transfalse
\begin{figure}[tb]
\begin{center}
\relsize{-0.5}
%\renewcommand{\arraystretch}{1.5}% Wider
\resizebox{\columnwidth}{!}{
    \begin{tabular}{|C{9.5cm}C{8cm}|} 
      \hline
  \begin{tabular}{R{9.5cm}}
  \RL{وأنت} $\stackrel{u1}{\framebox[1.2\width]{\RL{الأول}}}$ $\stackrel{p_2}{\framebox[1.2\width]{\RL{التقاطع}}}$ \RL{من} $\stackrel{r_1}{\framebox[1.2\width]{\RL{بالقرب}}}$ $\stackrel{n_1}{\framebox[1.2\width]{\RL{خليفة}}}$ $\stackrel{p_1}{\framebox[1.2\width]{\RL{برج}}}$ \RL{تلاحظ} \RL{ألا} \RL{المستحيل} \RL{من} 
  \\ ~ \\
	$\stackrel{u_2}{\framebox[1.2\width]{\RL{الأولى}}}$ \RL{المرّة} \RL{هذه} \RL{كانت} \RL{وإن} \RL{حتّى} $\stackrel{n_3}{\framebox[1.2\width]{\RL{زايد،}}}$ $\stackrel{n_2}{\framebox[1.2\width]{\RL{الشيخ}}}$ $\stackrel{p_3}{\framebox[1.2\width]{\RL{شارع}}}$ $\stackrel{r_2}{\framebox[1.2\width]{\RL{في}}}$ \RL{سيارتك} \RL{تقود} 
  \\ ~ \\
  $\stackrel{p_7}{\framebox[1.2\width]{\RL{المبنى}}}$ \RL{هذا} \RL{من} $\stackrel{r_4}{\framebox[1.2\width]{\RL{مقربة}}}$ \RL{على} $\stackrel{p_6}{\framebox[1.2\width]{\RL{مول}}}$ $\stackrel{p_5}{\framebox[1.2\width]{\RL{دبيّ}}}$ \RL{يقع} $\stackrel{p_4}{\framebox[1.2\width]{\RL{الطريق؛}}}$ \RL{هذا} $\stackrel{r_3}{\framebox[1.2\width]{\RL{فيها}}}$ \RL{تسلك} \RL{التي} 
  \\
  \RL{العالم.} \RL{في} \RL{الأطول} \RL{يُعد} \RL{الذي}  
  \\
  \multicolumn{1}{L{9.5cm}}{
   \arabfalse \transtrue
   \RL{من المستحيل ألا تلاحظ برج خليفة بالقرب من التقاطع الأول وأنت تقود سيارتك في شارع الشيخ زايد، حتّى وإن كانت هذه المرّة الأولى التي تسلك فيها هذا الطريق؛ يقع دبيّ مول على مقربة من هذا المبنى الذي يُعد الأطول في العالم.}
   \arabtrue \transfalse} 
  \\
  \multicolumn{1}{L{9.5cm}}{It is impossible not to notice $\stackrel{n1}{\framebox[1.2\width]{Khalifa}}$ $\stackrel{p_1}{\framebox[1.2\width]{Tower}}$ $\stackrel{r_1}{\framebox[1.2\width]{next to}}$ the $\stackrel{u_1}{\framebox[1.2\width]{first}}$ $\stackrel{p_2}{\framebox[1.2\width]{intersection}}$ while you are driving your car $\stackrel{r_2}{\framebox[1.2\width]{on}}$ $\stackrel{n_2}{\framebox[1.2\width]{Sheikh}}$ $\stackrel{n_3}{\framebox[1.2\width]{Zayed}}$ $\stackrel{p_3}{\framebox[1.2\width]{Road}}$, even if this was $\stackrel{u_2}{\framebox[1.2\width]{the first}}$ time that you take this $\stackrel{p_4}{\framebox[1.2\width]{road}}$; $\stackrel{p_5}{\framebox[1.2\width]{Dubai}}$ $\stackrel{p_6}{\framebox[1.2\width]{Mall}}$ is located $\stackrel{r_4}{\framebox[1.2\width]{near}}$ this $\stackrel{p_7}{\framebox[1.2\width]{building}}$, which is the longest in the world.} 
  \end{tabular}
&
\resizebox{0.6\columnwidth}{!}{
  \input{figures/ergraphEx.tex}}
 \\ \hline
%\multicolumn{1}{c}{\multirow{1}{*}{\textbf{(a) Text with directions}}} &
%\multicolumn{1}{c}{\textbf{(b) Formula, matches, and entity-relation graph}} \\
    \end{tabular}
}
\end{center}
  \caption{Direction example with Arabic text, annotated with entities, transliteration, translation, and extracted relational entities in a graph} 
  \label{fig:intromotiv}
\end{figure}
\transtrue
\setcode{standard}

>>>>

\transtrue
\setcode{utf8}
\setarab
\transfalse
\begin{figure}[tb]
\begin{center}
\relsize{-0.5}
%\renewcommand{\arraystretch}{1.5}% Wider
\resizebox{\columnwidth}{!}{
    \begin{tabular}{|C{10cm}C{8cm}|} 
      \hline
  \begin{tabular}{R{10cm}}
\RL{تقود} \RL{وأنت} $\stackrel{u1}{\framebox[1.2\width]{\RL{الأول}}}$ $\stackrel{p_2}{\framebox[1.2\width]{\RL{التقاطع}}}$ \RL{من} $\stackrel{r_1}{\framebox[1.2\width]{\RL{بالقرب}}}$ $\stackrel{n_1}{\framebox[1.2\width]{\RL{خليفة}}}$ $\stackrel{p_1}{\framebox[1.2\width]{\RL{برج}}}$ \RL{تلاحظ} \RL{ألا} \RL{المستحيل} \RL{من} 
  \\ ~ \\
$\stackrel{r_3}{\framebox[1.2\width]{\RL{فيها}}}$ \RL{تسلك} \RL{التي} $\stackrel{u_2}{\framebox[1.2\width]{\RL{الأولى}}}$ \RL{المرّة} \RL{هذه} \RL{كانت} \RL{وإن} \RL{حتّى} $\stackrel{n_3}{\framebox[1.2\width]{\RL{زايد،}}}$ $\stackrel{n_2}{\framebox[1.2\width]{\RL{الشيخ}}}$ $\stackrel{p_3}{\framebox[1.2\width]{\RL{شارع}}}$ $\stackrel{r_2}{\framebox[1.2\width]{\RL{في}}}$ \RL{سيارتك}
  \\ ~ \\
\RL{العالم.} \RL{في} \RL{الأطول} \RL{يُعد} \RL{الذي}  $\stackrel{p_7}{\framebox[1.2\width]{\RL{المبنى}}}$ \RL{هذا} \RL{من} $\stackrel{r_4}{\framebox[1.2\width]{\RL{مقربة}}}$ \RL{على} $\stackrel{p_6}{\framebox[1.2\width]{\RL{مول}}}$ $\stackrel{p_5}{\framebox[1.2\width]{\RL{دبيّ}}}$ \RL{يقع} $\stackrel{p_4}{\framebox[1.2\width]{\RL{الطريق؛}}}$ \RL{هذا}
  \\ ~ \\
%  \RL{العالم.} \RL{في} \RL{الأطول} \RL{يُعد} \RL{الذي}  
%  \\
  \multicolumn{1}{L{10cm}}{
   \arabfalse \transtrue
   \RL{من المستحيل ألا تلاحظ برج خليفة بالقرب من التقاطع الأول وأنت تقود سيارتك في شارع الشيخ زايد، حتّى وإن كانت هذه المرّة الأولى التي تسلك فيها هذا الطريق؛ يقع دبيّ مول على مقربة من هذا المبنى الذي يُعد الأطول في العالم.}
   \arabtrue \transfalse} 
  \\ ~ \\
  \multicolumn{1}{L{10cm}}{It is impossible not to notice $\stackrel{n1}{\framebox[1.2\width]{Khalifa}}$ $\stackrel{p_1}{\framebox[1.2\width]{Tower}}$ $\stackrel{r_1}{\framebox[1.2\width]{next to}}$ the $\stackrel{u_1}{\framebox[1.2\width]{first}}$ $\stackrel{p_2}{\framebox[1.2\width]{intersection}}$ while you are driving your car $\stackrel{r_2}{\framebox[1.2\width]{on}}$ $\stackrel{n_2}{\framebox[1.2\width]{Sheikh}}$ $\stackrel{n_3}{\framebox[1.2\width]{Zayed}}$ $\stackrel{p_3}{\framebox[1.2\width]{Road}}$, even if this was $\stackrel{u_2}{\framebox[1.2\width]{the first}}$ time that you take this $\stackrel{p_4}{\framebox[1.2\width]{road}}$; $\stackrel{p_5}{\framebox[1.2\width]{Dubai}}$ $\stackrel{p_6}{\framebox[1.2\width]{Mall}}$ is located $\stackrel{r_4}{\framebox[1.2\width]{near}}$ this $\stackrel{p_7}{\framebox[1.2\width]{building}}$, which is the longest in the world.} 
  \end{tabular}
&
\resizebox{0.6\columnwidth}{!}{
  \input{figures/ergraphEx.tex}}
\\  \hline
%\multicolumn{1}{c}{\multirow{1}{*}{\textbf{(a) Text with directions}}} &
%\multicolumn{1}{c}{\textbf{(b) Formula, matches, and entity-relation graph}} \\
    \end{tabular}
}
\end{center}
\vspace{-1em} 
  \caption{Direction example with Arabic text, annotated with entities, transliteration, translation, and extracted relational entities in a graph} 
  \label{fig:intromotiv}
\vspace{-1em} 
\end{figure}
\transtrue
\setcode{standard}

### 2018-06-23 tex/merf/TALMajor/gui.tex
format issues ficed: figure and table caption style
minor changes

\begin{figure}[tb!]
  \centering
  {\setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0.5pt}%
  \fbox{\includegraphics[width=\textwidth]{figures/mreedit}}}
  \caption{\framework tag type regular expression editor.}
  \label{f:sfe}
\end{figure}

@@

\vspace{-1em}
\subsection{Tag type regular expression editor}

@@

\begin{figure}[tb!]
  \centering
  \includegraphics[width=\textwidth]{figures/treegraph}
  \caption{MRE annotated Text, MRE \major{matching parse tree}, and entity-relation graph}
  \label{fig:treegraph}
\end{figure}

@@

The \major{matching parse tree} view shows the selected match in a graph view.
Figure~\ref{fig:treegraph} shows the \major{matching parse tree} of the direction task 
\RL{dby mwl `l_A mqrbT mn h_dA Almbn_A}(Dubai Mall is located near this building). 
%Leaf nodes show the match text, internal nodes show the operations, and edge labels shows the MBF tag type name.

@@

A snapshot of the GUI in 
Figure~\ref{fig:treegraph} shows in an interactive graph view
the entity-relation graph of the match of the user defined relation 
extracted from the \major{matching parse tree} of the MRE. 
%
In the computational action editor, an advanced user can 
enter C++ code and use the \framework API to program and process 
subexpression matches. 
%RE matches In order to add computational actions to an MRE sub-expression, 
%the user selects to edit the actions in the MRE tag type editor. 
%The view allows the user to specify the actions in a user friendly manner. 
%\framework provides the user with an API to access the sub-expression match features easily and define the pre-match actions. 
%The match features include the text, position, length, and numerical value if applicable, and morphological features for MBF matches. 
%\framework also enables the user to add C++ declarations and library includes.

>>>>

@@

\begin{figure}[tb]
  \centering
  {\setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0.5pt}%
  \fbox{\includegraphics[width=0.995\textwidth]{figures/mreedit}}}
  \vspace{-2em}
  \caption{\label{f:sfe}\framework tag type regular expression editor.}
\end{figure}

\subsection{Tag type regular expression editor}
\vspace{-1em}

@@

@@

The matching parse tree view shows the selected match in a graph view.
Figure~\ref{fig:treegraph} shows the matching parse tree of the direction task 
\RL{dby mwl `l_A mqrbT mn h_dA Almbn_A}(Dubai Mall is located near this building). 
%Leaf nodes show the match text, internal nodes show the operations, and edge labels shows the MBF tag type name.

\begin{figure}[tb]
  \centering
  \includegraphics[width=\textwidth]{figures/treegraph}
  \vspace{-2em}
  \caption{\label{fig:treegraph}MRE annotated Text, MRE matching parse tree, and entity-relation graph.}
\end{figure}

@@

A snapshot of the GUI in 
Figure~\ref{fig:treegraph} shows in an interactive graph view
the entity-relation graph of the match of the user defined relation 
extracted from the matching parse tree of the MRE. 
%
In the computational action editor, an advanced user can 
enter C++ code and use the \framework API to program and process 
subexpression matches. 
%RE matches In order to add computational actions to an MRE sub-expression, 
%the user selects to edit the actions in the MRE tag type editor. 
%The view allows the user to specify the actions in a user friendly manner. 
%\framework provides the user with an API to access the sub-expression match features easily and define the pre-match actions. 
%The match features include the text, position, length, and numerical value if applicable, and morphological features for MBF matches. 
%\framework also enables the user to add C++ declarations and library includes.

### 2018-06-23 tex/merf/TALMajor/introduction.tex
format issues ficed: figure and table caption style
minor changes

\major{
Figure~\ref{fig:intromotiv} illustrates the target of \framework using 
the directions to Dubai Mall example~\footnote{text taken from the Dubai 
Mall website ~\url{http://www.thedubaimall.com/ar/}}.
The figure also presents a transliteration and an English translation of 
the Arabic text.
}
The framed words in the text are entities referring to names of people 
($n_1,n_2,n_3$), 
names of places ($p_1,\dots,p_7$), 
relative positions ($r_1,\dots,r_4$), 
and numerical terms ($u_1,u_2$). 
We would like to extract those entities, and then extract
the relational entities forming the graph in Figure~\ref{fig:intromotiv} 
where vertices express entities, 
and edges represent the relational entities.

@@

In this paper, 
we present \framework, a morphology-based entity and relational entity
extraction framework for Arabic text.
\framework provides a user-friendly interface where the user defines tag types 
and associates them with
%\framework formulae that are 
regular expressions over 
%\framework
Boolean formulae.
%
A Boolean formula is defined by a term, negation of a term, or disjunction of terms.
Terms are matches to Arabic morphological features including 
prefix, stem, suffix, part of speech (POS) tags, gloss tags, extended synonym 
tags, and semantic categories.
\major{
For example, entity $p_1$ in Figure~\ref{fig:intromotiv} has a place semantic 
category. 
}
\framework regular expressions support operators such as concatenation, 
zero or one, zero or more, one or more, up to $M$ \major{repititions where $M$ is 
a non-zero positive integer}, and logical conjunction and disjunction.
\major{
For example, the sequence between $p_1$ and $p_2$ matches a regular expression $re$ 
that requires two semantic place categories with a place-preposition POS 
tag ($r_1$) in between. 
}

An editor allows the user to associate an action with each sub-expression.
The user specifies the action with C++ code and uses an API to access information 
related to the matches such as text, position, length, morphological features, 
and numerical value.
%
\major{
Each regular expression is associated with a named identifier to form a 
{\em local grammar} like structure~\cite{traboulsi2009arabic}. 
%
A relation definition GUI allows the user to provide relational tuples 
where each tuple has a 
source, a destination and an edge label.
The user uses the regular expression identifiers to define the relational 
tuple elements.
For example, the relation between $e_1, e_2$ and $r$ shown in 
Figure~\ref{fig:intromotiv} is a match of a relational tuple 
over the components of $re$. 
%
We refer to regular expressions and Boolean formulae as expressions and formulae, respectively.
We also refer to expressions as rules
when used in a grammar context; e.g.  when used with an identifier. 
%We discuss the importance of the morphological features supported in 
%\framework terms in Section~\ref{sec:morph}; in brief, morphological preprocessing
%is key to Arabic NLP.% due to the richness of Arabic morphology.
}

@@

In \framework, we make the following contributions.
\begin{itemize}
  \item \framework enables the user to define relations in a simple manner
    and automatically detects relational entities matching the user defined relations. 
  \item \framework enables the user to associate sub-expressions
    with code actions, and executes the code action 
    when a corresponding match is found.
    It also provides an API to enable access to match 
    features such as text, position, length, numerical value, and morphological features.
  \item \framework enables the user to tag words based on a 
    \major{
    novel ligh Arabic wordnet relation that leverages the 
    synonym $Syn^k$ feature.
  }

  \item \major{
\framework is open source and available online for the research community 
    under on (\url{https://github.com/codelogicanalysis/atmine}).}
\end{itemize}

>>>>

Figure~\ref{fig:intromotiv} illustrates the target of \framework using 
the directions to Dubai Mall example~\footnote{text taken from the Dubai 
Mall website ~\url{http://www.thedubaimall.com/ar/}}.
The figure also presents a transliteration and an English translation of 
the Arabic text.
The framed words in the text are entities referring to names of people 
($n_1,n_2,n_3$), 
names of places ($p_1,\dots,p_7$), 
relative positions ($r_1,\dots,r_4$), 
and numerical terms ($u_1,u_2$). 
We would like to extract those entities, and then extract
the relational entities forming the graph in Figure~\ref{fig:intromotiv} 
where vertices express entities, 
and edges represent the relational entities.

@@

In this paper, 
we present \framework, a morphology-based entity and relational entity
extraction framework for Arabic text.
\framework provides a user-friendly interface where the user defines tag types 
and associates them with
%\framework formulae that are 
regular expressions over 
%\framework
Boolean formulae.
%
A Boolean formula is defined by a term, negation of a term, or disjunction of terms.
Terms are matches to Arabic morphological features including 
prefix, stem, suffix, part of speech (POS) tags, gloss tags, extended synonym 
tags, and semantic categories.
For example, entity $p_1$ in Figure~\ref{fig:intromotiv} has a place semantic 
category.
\framework regular expressions support operators such as concatenation, 
zero or one, zero or more, one or more, up to $M$ repetitions where $M$ is 
a non-zero positive integer, and logical conjunction and disjunction.
For example, the sequence between $p_1$ and $p_2$ matches a regular expression $re$ 
that requires two semantic place categories with a place-preposition POS 
tag ($r_1$) in between.

An editor allows the user to associate an action with each sub-expression.
The user specifies the action with C++ code and uses an API to access information 
related to the matches such as text, position, length, morphological features, 
and numerical value.
%
Each regular expression is associated with a named identifier to form a 
{\em local grammar} like structure~\cite{traboulsi2009arabic}. 
%
A relation definition GUI allows the user to provide relational tuples 
where each tuple has a 
source, a destination and an edge label.
The user uses the regular expression identifiers to define the relational 
tuple elements.
For example, the relation between $e_1, e_2$ and $r$ shown in 
Figure~\ref{fig:intromotiv} is a match of a relational tuple 
over the components of $re$. 
%
We refer to regular expressions and Boolean formulae as expressions and formulae, respectively.
We also refer to expressions as rules
when used in a grammar context; e.g.  when used with an identifier. 
%We discuss the importance of the morphological features supported in 
%\framework terms in Section~\ref{sec:morph}; in brief, morphological preprocessing
%is key to Arabic NLP.% due to the richness of Arabic morphology.

@@

In \framework, we make the following contributions.
\begin{itemize}
  \item \framework enables the user to define relations in a simple manner
    and automatically detects relational entities matching the user defined relations. 
  \item \framework enables the user to associate sub-expressions
    with code actions, and executes the code action 
    when a corresponding match is found.
    It also provides an API to enable access to match 
    features such as text, position, length, numerical value, and morphological features.
  \item \framework enables the user to tag words based on a novel light Arabic wordnet relation that leverages the synonym $Syn^k$ feature.
  \item \framework is open source and available online for the research community 
    under on (\url{https://github.com/codelogicanalysis/atmine}).
\end{itemize}

### 2018-06-23 tex/merf/TALMajor/relatedwork.tex
minor changes

\major{
Another track in the literature targets specific tasks such as NER using 
statistical and machine learning techniques such as maximum entropy, 
optimized feature sets and conditional random fields~\cite{benajiba2007anersys,benajiba2008arabic,ekbal2010named,abdelrahman2010integrated}.
}
%The work in~\cite{ekbal2010named} presents a language independent approach for NER extraction using {\em support vector machines}. 
%The work in~\cite{abdelrahman2010integrated} integrates a semi-supervised bootstrapping pattern recognition technique, and a supervised classifier based on {\em conditional random fields} to solve NER problems.
Knowledge-based techniques such 
as~\cite{zaghouani2010adapting,traboulsi2009arabic} propose local grammars with morphological stemming.% to perform NER. 
%The work in~
\cite{ZaMaHaCicling2012Entity} present a method for extracting entities and events, and relations among them from Arabic text using a hierarchy of manually built 
finite state machines driven by morphological features, 
and graph transformation algorithms. 
Such techniques require advanced linguistic and programming expertise.

>>>>

Another track in the literature targets specific tasks such as NER using 
statistical and machine learning techniques such as maximum entropy, 
optimized feature sets and conditional random fields~\cite{benajiba2007anersys,benajiba2008arabic,ekbal2010named,abdelrahman2010integrated}.
%The work in~\cite{ekbal2010named} presents a language independent approach for NER extraction using {\em support vector machines}. 
%The work in~\cite{abdelrahman2010integrated} integrates a semi-supervised bootstrapping pattern recognition technique, and a supervised classifier based on {\em conditional random fields} to solve NER problems.
Knowledge-based techniques such 
as~\cite{zaghouani2010adapting,traboulsi2009arabic} propose local grammars with morphological stemming.% to perform NER. 
%The work in~
\cite{ZaMaHaCicling2012Entity} present a method for extracting entities and events, and relations among them from Arabic text using a hierarchy of manually built 
finite state machines driven by morphological features, 
and graph transformation algorithms. 
Such techniques require advanced linguistic and programming expertise.
